---
title: "The Impact of Vaccines"
code name : "Project Vax"
author: "project team members"
Winter 2022
date: "2/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**Authors:**
Emma Bacarra - ebacarra@uw.edu
Jayden Set - jset@uw.edu
Larry Nguyen - nguyenla@uw.edu
Parker - parker10@uw.edu

**Affiliation:** INFO-201: Technical Foundations of Informatics - The Information School - University of Washington

### Abstract
Our main question is how vaccines can create an impact on public health in the U.S. This question is important because vaccinations can lower an individual’s risk
for getting or spreading the virus, prevent severe symptoms, illness, and death, and initiates an antibody response in the immune system. To address the question,
we will compare vaccination rates to the mortality rate and the total number of cases.

### Keywords
Vaccinations, spread prevention, public health, and mortality rate.

### 1.0 Introduction
The dataset(s) we are analyzing shows us the rate of vaccinations across the United States and the infection/mortality rate among those vaccinated and unvaccinated
due to COVID-19. Our group is interested in exploring the effectiveness of the COVID-19 vaccines and how the data either supports or opposes the current iterations
of the COVID-19 vaccines in the fight against the virus and its variants. The outbreak of the Coronavirus vaccine has greatly disrupted modern life the past few
years and these vaccines can be the key to normality within the near future.

### 2.0 Design Situation
Our group’s topic of concern revolves around the COVID-19 and its relationship between vaccinated and unvaccinated people, and how vaccinations have a role to play
in the mortality rate which has been surging since the pandemic has started. The concern among getting everyone vaccinated stems from the human values of mutual
empathy and the love for your common man. We want everyone to stay healthy and alive. There is also the selfish concern of getting life back to normal, before the
global pandemic and the restrictions that came with it. The goal of vaccines is to reach herd immunity. In order to protect those vulnerable and unable to be
vaccinated, we strive to get every abled body vaccinated. In order to protect those vulnerable and unable to be vaccinated, we strive to get every abled body
vaccinated. The direct stakeholders, those who will directly interact with **Project Vax**, are political leaders and policymakers, whether local or federal, who
can explain the effectiveness of vaccines. The indirect stakeholders are those who did not directly interact with the project firsthand but can feel the effects
from those who did. Indirect stakeholders may include the general public who are subject to laws and regulations placed upon them by policy makers and government
officials who have directly interacted with **Project Vax**. They may also include individuals who are at risk of contracting COVID-19 and unable to vaccinate
themselves due to preexisting conditions or individuals who refuse to be vaccinated and are being educated by those who are.  Lastly, these may include essential
workers who are subject to regulations that may require them to be vaccinated or risk losing their jobs. Essential workers are put at risk of exposure and close
contact in their jobs everyday, and are in even more danger if the individuals they interact with are not doing their part to be cautious. The harms of **Project
Vax**  are the limitations of the existing data and its accuracy, the information can be biased to service political motivations that are promptly against
vaccines. However, the benefits of our project heavily outweigh the harms. With the information present in **Project Vax**; national, state and local governments
are better able to assess the situation of the global pandemic and enact policies and regulations accordingly. The hope is for representatives in the United
States’ lowest vaccinated states to enact change as deaths from the virus are highest among low vaccinated states (NBCNews, 2022). This can also include the
expansion of free or affordable COVID-19 vaccines that are made accessible to the general public.

### 3.0 Research Questions
Our teams questions:
- How much of an impact do vaccines create on public health in the U.S.?
- Have vaccinations helped the mortality rate of COVID-19 go down?
- Are a majority of the population in countries fully vaxxed?
- As the amount of new deaths were rising, were more people willing to get vaccinated?

These questions are important because they are focused and aimed to find the positives that come with becoming vaccinated in order to prevent COVID-19’s death toll
from continuing to go up. With these questions, we want to be able to find the positives of being vaccinated and prove that it is something that could eventually
take down the virus. We also would like to shed light on how the vaccination actually works and could prevent death, doing this we could potentially motivate
people to go get vaccinated in order to combat the virus. We will address these questions by using our [dataset on vaccination vs death]
(https://www.kaggle.com/sinakaraji/covid-vaccination-vs-death) . By doing so we could look at all countries across the globe and view content such as how many
people have been fully vaccinated and the ratio comparing the total population to the number of people who have got vaccinated.

### 4.0 Data Set
The dataset we have chosen to use for our project _The Impact of Vaccines_, compares the [covid-19 vaccination rate vs the mortality rate across the world]
(https://www.kaggle.com/sinakaraji/covid-vaccination-vs-death). This set of data has 15,803 views, 2,758 downloads, and 34 contributors. The top contributing
countries are South Korea, China, Iran, and Russia. It observes the outbreak that we have all witnessed and looks at the positives around the vaccination that has
surfaced. The reason why this dataset was generated was to investigate the impact of the coronavirus vaccine on the covid-19 mortality. This data file exceeds the
amount of 200 observations and is a combination of a total of 3 different datasets which are sourced from Kaggle and the World Health Organization. It represents a
majority of each country around the world and keeps track of their total vaccinations, number of people vaccinated, people fully vaccinated, new deaths per day,
the population count, and the ratio of vaccinated individuals per available vaccinations. The majority of countries had a ratio of 0 -  6.11. The most common
country to appear in this data set is the United States. The data was collected by Sina Karaji and it was updated last about a month ago meaning that it is up to
date. The expected updated frequency of this dataset is every month. The reason why he created the dataset was because he noticed how many deaths have been caused
since the outbreak of the coronavirus, so he wanted to look at the impact that the vaccination has had and how it has contributed to the progress beside global
death counts. The author is also motivated to provide this information because of anti-vaccine groups that form even though vaccines have been proven to have many
benefits. By doing this he is able to show people the positive impact of the vaccination, persuade the rest of the world to get vaccinated, and positively
influence the impact of coronavirus vaccinations on coronavirus mentality. Those who are likely to be benefited from the data could be companies such as the CDC
who are looking to view datasets that could help them combat the virus, but this could also interest people who are trying to convince anti-vaxxers that
vaccinations are helpful and the public who wants to see the positive effect that the vaccine has had.

### 5.0 Expected Implications
Our group expects that local, state, and federal policymakers will gain a greater understanding of the effectiveness of COVID-19 vaccines from our project “The
Impact of Vaccines”. The goal of our project is to inspire those will the power and influence to change policy/ enact laws that will ensure their citizens have
access to free or affordable vaccines in order to reach herd immunity (which is when a large portion of the population becomes immune to a disease, it makes the
spread of the disease from one individual to another very unlikely. As a result, the whole community is protected). This can be done through utilizing the large
size of government to acquire vaccines in bulk quantities at a low cost. Governments can also utilize data science to strategically move vaccines to cities and
states where demand and need are greatest. Not only do vaccines have to be financially accessible to everyone but it should be easy to find a vaccination site
locally and schedule an appointment with ease. We also expect **Project Vax** will be used as an educational tool to convince those who are resistant to receiving
a vaccine or those who do not believe in the effectiveness of vaccines.

### 6.0 Limitations
Our group acknowledges that there are certain limitations to this research project, such as
access to the most current and accurate information. Because of the scale of our research project, we are limited to data provided by third party researchers who
publish their records online. Moreover, the research from these studies may not be inclusive of all demographics in an entire population;  in order to classify
that an individual has been vaccinated or tested positive for COVID-19, they had to have access to a testing site and proper healthcare provisions. In the United
States, there are certain demographics in our population (especially low-income communities of color) that do not have access to proper provisions, which means
that the data we find on the internet won’t include individuals in those communities. There are additional factors such as the vaccine manufacturer, whether it is
Pfizer, Moderna, or Johnson & Johnson or the different strains of COVID-19 (Delta, Omicron). Our research project also classifies a vaccinated individual as anyone
who received at least one dose of the vaccine. These factors may influence the infection and mortality rate of our research project. We must also take account of
factors that contribute to the number of vaccinated or unvaccinated individuals in a population, such as medical exemptions due to allergies; however, the CDC has
explicitly stated that for those with underlying medical conditions, the COVID-19 vaccination is safe to receive.

### References
Hanage, William P., et al. “COVID-19: US Federal Accountability for Entry, Spread, and Inequities—Lessons for the Future.” European Journal of Epidemiology, vol.
35, no. 11, Nov. 2020, pp. 995–1006, link.springer.com/article/10.1007/s10654-020-00689-2#citeas, 10.1007/s10654-020-00689-2.

Bove, Tristan. “Fully Vaccinated People Are 93% Less Likely to Die of COVID Compared to Unvaccinated People.” Fortune, Fortune, 4 Feb. 2022,
fortune.com/2022/02/04/fully-vaccinated-93-percent-less-likely-covid-death-compared-unvaccinated/.

Chiwaya, Nigel. “900,000 Dead: Covid Deaths Are Surging in Low-Vaccination States.” NBC News, NBC News, 4 Feb. 2022, www.nbcnews.com/news/us-news/900-000-dead-
covid-deaths-are-surging-low-vaccination-states-n1288586.

## What is this file for?

* This file will eventually become your project report for **P02: Exploratory Data Analysis**. Specifically, you will write rmarkdown to report your exploratory data analysis.

* Please see Canvas for more details.

## ACCESSING THE /source DIRECTORY
### Example 1: Source some code and run a function
```{r}
# Example 1: Note relative path, which can be read: Up one
# directory(..), down into source (/source), and
# then "source" an R file (data_access.R)
source("../source/data_access.R")
data_access_test()
```

### Aggregate Table
```{r}
library(dplyr)

data <- read.csv("https://github.com/info-201a-wi22/final-project-starter-info/raw/main/data/covid-vaccination-vs-death_ratio.csv")

View(data)

earlycountries <-
  data %>%
  group_by(country) %>%
  filter(date == min(date)) %>%
  summarize(
    country,
    date,
    total_vaccinations,
    New_deaths,
    ratio
    ) %>%
  rename(early_date = date)
View(earlycountries)

latecountries <-
  data %>%
  group_by(country) %>%
  filter(date == max(date)) %>%
  summarize(
    country,
    date,
    total_vaccinations,
    New_deaths,
    ratio
  ) %>%
  rename(recent_date = date)
View(latecountries)

compare <-
  left_join(latecountries, earlycountries, by = "country") %>%
  mutate(
    time_passed = as.Date(compare$recent_date) - as.Date(compare$early_date),
    ratio_diff = compare$ratio.x - compare$ratio.y,
    vax_change = compare$total_vaccinations.x - compare$total_vaccinations.y,
    death_change = compare$New_deaths.x - compare$New_deaths.y
    ) %>%
  select(
    country,
    early_date,
    recent_date,
    time_passed,
    vax_change,
    death_change,
    ratio_diff
    )
View(compare)

us_info <- compare[(compare$country == "United States of America"), ]
us_info
```
Our aggregate table shows us the most the earliest data on total vaccinations per country and ratio between vaccinations and deaths. We then summarize the data
for the most recent dates per country and its ratio between total vaccinations and deaths. Our goal is to showcase that the as more and more people become
vaccinated, the ratio between total vaccinations and new deaths would drop exponentally.

### Chart 1
```{r}
library("tidyverse")
library("ggplot2")

dataset <- read.csv("https://raw.githubusercontent.com/info-201a-wi22/final-project-starter-info/main/data/covid-vaccination-vs-death_ratio.csv")

dataset %>%
  filter(country == "United States of America") %>%
  ggplot(aes(x = date, y = total_vaccinations, colour = country))+
  geom_point(size = 5, alpha = 0.2)+
  geom_line(size = 1)+
  theme_minimal()+
  labs(title = "Progession of Vaccinations in the US")

#This line chart shows the progessioin of vaccinations within the US and how they have increased with time and how long COVID-19 has been around.

### Chart 2
library(dplyr)
library(ggplot2)

data <- read.csv("https://data.cdc.gov/api/views/8xkx-amqh/rows.csv?accessType=DOWNLOAD&bom=true&format=true")
View(data)

colnames(data)[1] <- "Date"
data$date <- as.Date(data$Date, "%m/%d/%Y")
data$date
class(data$date)

recentdata <-
  data %>%
  filter(date == (max(date))) %>%
  summarize(
    Date,
    Recip_County,
    Recip_State,
    Census2019, # census population
    Administered_Dose1_Pop_Pct, # % w/ at least 1 dose by state
    Census2019_5PlusPop, # census population age group 5
    Administered_Dose1_Recip_5PlusPop_Pct, # % at least 1 dose age group 5
    Census2019_12PlusPop, # census population age group 12
    Administered_Dose1_Recip_12PlusPop_Pct, # % at least 1 dose age group 12
    Census2019_18PlusPop, # census population age group 18
    Administered_Dose1_Recip_18PlusPop_Pct, # % at least 1 dose age group 18
    Census2019_65PlusPop, # census population age group 65
    Administered_Dose1_Recip_65PlusPop_Pct # % at least 1 dose age group 65
  ) %>%
  rename(
    county = Recip_County,
    state = Recip_State,
    totpop_2019 = Census2019,
    state_vax_pct = Administered_Dose1_Pop_Pct,
    age5_pop2019 = Census2019_5PlusPop,
    age5_pctvax = Administered_Dose1_Recip_5PlusPop_Pct,
    age12_pop2019 = Census2019_12PlusPop,
    age12_pctvax = Administered_Dose1_Recip_12PlusPop_Pct,
    age18_pop2019 = Census2019_18PlusPop,
    age18_pctvax = Administered_Dose1_Recip_18PlusPop_Pct,
    age65_pop2019 = Census2019_65PlusPop,
    age65_pctvax = Administered_Dose1_Recip_65PlusPop_Pct
  )
View(recentdata)

# `recentdata` compares population of a county and percentage of that population
# with at least 1 dose of the vaccine. Includes general population (i.e. consensus
# population) and general vaccinated percentage, along with statistics by age groups:
# 5+, 12+, 18+, and 65+.

wa_state_data <-
  recentdata %>%
  filter(state == "WA")
View(wa_state_data)

ggplot(wa_state_data, aes(x=county, y=state_vax_pct)) +
  geom_point(size=2, shape=20)

# This scatter plot compares counties in WA state to the percentages of those county
# populations with at least 1 dose of the vaccine. Data is pulled from the most
# recent date where vaccinations have been recorded.
```
